\Large \bf Vigor: Toward Carrier-Grade Software Network Functions

### 2016-12-01 doc/paper/paper.tex
Restructured second section according to the last project meeting.

\subsection{Developer experience}

@@

  while(1) {^\label{line:loop-begin}^
    if (queue_len < CAP) {
      int end = (begin + queue_len)%CAP;
      if (receive_packet(&packets[end]) && ^\label{line:recv-pkt}^
          packets[end].port != 9)
        queue_len += 1;
    }
    if (0 < queue_len && can_send_packet()) { ^\label{line:can-send}^
      send_packet(&packets[begin]); ^\label{line:send-pkt}^
      begin = (begin + 1)%CAP;
    }
  }^\label{line:loop-end}^
  return 0;
}
\end{lstlisting}
  \caption{Original implementation of the Drop protocol. Contains an infinite
    loop (ll.\ref{line:loop-begin}-\ref{line:loop-end}) and a latent ring
    implementation (declared on l.\ref{line:pkt-def},\ref{line:iter-def})}
  \label{lst:orig}
\end{figure}

@@

@@

\subsection{Under the hood}
\label{sec:under-the-hood}

@@

Vigor transforms infinite loops using loop
invariants~\cite[\S~2.1]{cormen2009introduction} that allow to model all the
iterations of a loop in a single one. To support this kind of generalization, we
added automatic havoc-ing~\cite{barnett2005boogie} to KLEE. Havoc-ing simply
replaces all the variables that may change in the loop by unconstrained symbols
at the beginning of the iteration. In our example that would be the variables
\code{begin} and \code{queue_len} and the contents of \code{packets[0..CAP]},
but not the value of the \code{packets} pointer because it never changes. The
step 1 enables Vigor to reduce each problematic loop to a single generalized
iteration that allows reasoning in one pass.

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. Another classic SE challenge ---
``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset} spoils
the party here. When indexing an array with a symbolic variable (or when
dereferencing a symbolic pointer), the constraints on program state become
extremely complex, because the symbolic pointer or index could be referencing
many different possible locations. The complexity of the constraints chokes the
SMT solver, and the SEE grinds to a halt. In general, SE is not practical for
code with symbolic pointers or offsets. Alas, stateful C code almost always has
some of these. In this example the symbolic indexing comes from the ring array.
Once we replaced the infinite loop with a single iteration we have to assume
(due to havoc-ing) the \code{begin} and \code{begin + queue_len} indexes may
point anywhere in the array.

In Vigor, we circumvent the problem of symbolic indexing by encapsulating
program state into well-defined data structures and putting them into a
``library of components'' (e.g. the ring in step 2). This works well for NF
development, because the NF code uses a relatively small set of standard
structures, such as ring buffers, hash tables, least-prefix-match tables, etc.
We provide to the SEE symbolic models of these abstracted components that
``hide'' symbolic indexes and pointers from the SEE.

After the step 2 exhaustive SE takes less than 100 millisec regardless of the
value of \code{CAP}. The fact that the code does not crash on any path amounts
to a proof of crash freedom. Or does it? The validity of the proof depends on
our library implementation of \code{ring} not crashing; and the symbolic model
of \code{ring} accurately capturing how the \code{ring} implementation works.

The Vigor library comes with a formal specification and a formal proof that the
implementation satisfies that specification. In other words, there is a formal
spec for \code{ring} and a proof that the \code{ring} implementation satisfies
the spec. This is how we enable SE to verify the high level code, while using
separation logic to verify the library of components --- two different
verifications techniques for two different parts of the complete application.
Doing the library verification requires deep expertise in program verification,
but our effort is amortized across all NFs that use the library.

@@

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  ASSUME(packet_constraints(p));^\label{line:assume}^
}
\end{lstlisting}
  \caption{A symbolic model for \code{ring_pop_front}. It fills symbolically a
    packet and adds a constraint to it.}
  \label{lst:symbolic-model}
\end{figure}

@@

@@

The formal verification engine checks this sequence using contracts like the one
shown in the Fig.~\ref{lst:contract}, and ensures the abscence of misuse e.g.
call for \code{ring_pop} on a possibly empty ring. It then checks that the model
output covers all the possible outputs specified in the contract for the last
function. This happens in the last line on the listing. The assert contains the
path-constraint generated by the symbolic model. In this case the contract
(Fig.\ref{lst:contract}) guarantees that the popped packet satisfies
\code{user_property} that entails \code{arg2.port != 9}, hence the trace is
legitimate. (see~\S\ref{sec:model-validation-criteria}).

Let us follow now the `port != 9' condition across the framework. During
exhaustive SE, SEE first learns the constraint in the \code{if} condition:
Fig.~\ref{lst:vigor} l.\ref{line:if-cond}. Here, SEE just records it in the call
sequence for model validation, and does not use further in the SE, because our
model of \code{ring_push_back} discards the packet being pushed. However, the
call to \code{ring_pop_front} (l.\ref{line:pop-front}) reintroduces this
constraint formulated in \code{packet_constraints}
(Fig.~\ref{lst:customizations},l.\ref{line:c-pkt-cnstr}), according to the
symbolic model - Fig.~\ref{lst:symbolic-model}. Being applied to the fresh
packet from the ring, it satisfies the \code{VIGOR_CHECK} (Fig.~\ref{lst:vigor}
l.\ref{line:vigcheck}).

@@

\subsection{Detecting model inconsistencies}

The component developer produces a number of model that capture the component
semantics in more or less details for different usecases. For a particular use
case --- a NF, a particular model may be inappropriate. An inappropriate model
may not compromise the verification result, because Vigor detects both
overapproximate and underapproximate models.

@@

Suppose now we had an underapproximate model: Fig.~\ref{lst:underapproximate}.
Then the last line in the call sequence check (Fig.\ref{lst:call-sequence}) in
this case would look different:
\begin{lstlisting}
/*@ assert(arg2.port == 0); @*/
\end{lstlisting}
The assertion fails, because the contract (Fig.~\ref{lst:contract}) specifies
a wider range for \code{arg2.port} than just \code{0}.

Different models may work well for different usecases. E.g. the overapproximate
model is just right for a buffer, that does not drop the packets with port 9.
The underapproximate model would be just right for a buffer that forwards only the
packets with port equeal to 0.

>>>>

\subsection{Developer Experience}

@@

  while(1) {^\label{line:loop-begin}^
    if (queue_len < CAP) {
      int end = (begin + queue_len)%CAP;
      if (receive_packet(&packets[end]) && ^\label{line:recv-pkt}^
          packets[end].port != 9)
        queue_len += 1;
    }
    if (0 < queue_len && can_send_packet()) { ^\label{line:can-send}^
      send_packet(&packets[begin]); ^\label{line:send-pkt}^
      begin = (begin + 1)%CAP;
    }
  }^\label{line:loop-end}^
  return 0;
}
\end{lstlisting}
  \caption{Original implementation of the Drop protocol. Contains an infinite
    loop (ll.\ref{line:loop-begin}-\ref{line:loop-end}) and a latent ring
    implementation (declared on ll.\ref{line:pkt-def},\ref{line:iter-def})}
  \label{lst:orig}
\end{figure}

@@

Vigor provides two things: a verified library of components and a verification
toolchain. Each component consists of an API with formal specification, verified
implementation and a set of symbolic models. The toolchain verifies the NF code
and stitches it with the relevant components to produce a single reliable
application.

@@

\subsection{Under The Hood}
\label{sec:under-the-hood}

\subsubsection{Symbolic Execution}

@@

Vigor transforms infinite loops using loop
invariants~\cite[\S~2.1]{cormen2009introduction} that allow to model all the
iterations of a loop in a single one. To support this, we added automatic
havoc-ing~\cite{barnett2005boogie} to KLEE. Havoc-ing simply replaces all the
variables that may change in the loop by unconstrained symbols at the beginning
of the iteration. In our example that would be the variables \code{begin} and
\code{queue_len} and the contents of \code{packets[0..CAP]}, but not the value
of the \code{packets} pointer itself because it never changes. Step 1 enables Vigor
to reduce each problematic loop to a single generalized iteration that allows
reasoning in one pass.

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. Another classic SE challenge ---
``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset} ---
spoils the party here. When indexing an array with a symbolic variable (or when
dereferencing a symbolic pointer), the constraints on program state become
extremely complex, because the symbolic pointer or index could be referencing
many different possible locations. The complexity of the constraints chokes the
SMT solver, and the SEE grinds to a halt. In general, SE is not practical for
code with symbolic pointers or offsets. Alas, stateful C code almost always has
some of these. In this example the symbolic indexing comes from the ring array.
Once we replaced the infinite loop with a single iteration we have to assume
(due to havoc-ing) the \code{begin} and \code{begin + queue_len} indexes may
point anywhere in the array.

In Vigor, we circumvent the problem of symbolic indexing by encapsulating
program state into well-defined data structures and putting them into a
``library of components'' (e.g. the ring in step 2). This works well for NF
development, because the NF code uses a relatively small set of standard
structures, such as ring buffers, hash tables, least-prefix-match tables, etc.
We provide to the SEE symbolic models of these abstracted components that
``hide'' symbolic indexes and pointers from the SEE. After the step 2 exhaustive
SE takes less than 100 millisec regardless of the value of \code{CAP}.

SE is a first step on the way of a complete proof. After the exhaustive SE found
no crashes in the application code, we need to verify for each component (e.g.
for the ring component) that (1) the actual component implementation does not
crash itself and follows the declared API specification, and (2) the model used
during SE faithfully approximated the same specification.

\subsubsection{Formal Verification}

The Vigor library comes with a formal specification and a formal proof that the
implementation satisfies that specification. In other words, there is a formal
spec for \code{ring} and a proof that the \code{ring} implementation satisfies
the spec. This is how we enable SE to verify the NF code, while using
separation logic to verify the library of components --- two different
verification techniques for two different parts of the complete application.
Doing the library verification requires deep expertise in program verification,
but our effort is amortized across all NFs that use the library.

@@

\subsubsection{Symbolic Models}

The real implementation, even the correct one will inevitably complicate SE
of the NF code bringing back all the problems that we were so carefully
avoiding. To make the exhaustive SE feasible, Vigor uses symbolic models,
simplified implementation with small state that approximately mimics the real
thing in most cases.

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet), ^\label{line:fill-symbolic}^
                "popped_packet");
  ASSUME(packet_constraints(p));^\label{line:assume}^
}
\end{lstlisting}
  \caption{A symbolic model for \code{ring_pop_front}. It fills symbolically a
    packet (l.\ref{line:fill-symbolic}) and adds a constraint to it
    (l.\ref{line:assume}).}
  \label{lst:symbolic-model}
\end{figure}

@@

\subsubsection{Model Validation}

@@

The formal verification engine checks this sequence using contracts like the one
shown in the Fig.~\ref{lst:contract}, and ensures the abscence of misuse e.g.
call for \code{ring_pop} on a possibly empty ring. It then checks that the model
output covers all the possible outputs specified in the contract for the last
function. This happens on the line~\ref{line:assert}. The assert contains the
path-constraint generated by the symbolic model. In this case the contract
(Fig.\ref{lst:contract}) guarantees that the popped packet satisfies
\code{user_property} that entails \code{arg2.port != 9}, hence the trace is
legitimate. (see~\S\ref{sec:model-validation-criteria}).

Let us follow now the `port != 9' condition across the framework. During
exhaustive SE, SEE first learns the constraint in the \code{if} condition:
Fig.~\ref{lst:vigor} l.\ref{line:if-cond}. Here, SEE just records it in the call
sequence for model validation, and does not use further in the SE, because our
model of \code{ring_push_back} discards the packet being pushed. However, the
call to \code{ring_pop_front} (l.\ref{line:pop-front}) reintroduces this
constraint formulated in \code{packet_constraints}
(Fig.~\ref{lst:customizations}~l.\ref{line:c-pkt-cnstr}), according to the
symbolic model - Fig.~\ref{lst:symbolic-model}. Being applied to the fresh
packet from the ring, it satisfies the \code{VIGOR_CHECK}
(Fig.~\ref{lst:vigor}~l.\ref{line:vigcheck}).

@@

\subsubsection{Detecting model inconsistencies}

The component developer produces several model that capture the component
semantics in more or less details for different usecases. For a particular use
case --- a NF, a particular model may be inappropriate. An inappropriate model
may not compromise the verification result, because Vigor detects both
over- and under-approximations.

@@

Suppose now we had an underapproximate model: Fig.~\ref{lst:underapproximate}.
Then the last line in the call sequence check (Fig.\ref{lst:call-sequence}) in
this case would look different:
\begin{lstlisting}
/*@ assert(arg2.port == 0); @*/
\end{lstlisting}
Vigor model validation phase flags this inconsistency. Specifically, the
assertion fails, because the contract (Fig.~\ref{lst:contract}) specifies a
wider range for \code{arg2.port} than just \code{0}.

Different models may work well for different usecases. E.g. the overapproximate
model is just right for a buffer, that does not drop the packets with port 9.
The underapproximate model would be just right for a buffer that forwards only the
packets with port equal to 0.
